Introducing ICBe: An Event Extraction Dataset from Narratives about International Crises

### 2023-11-20 replication_paper/psrm_draft/ICBe_manuscript.Rmd
fig 2 footnote future research

We find substantive variation in recall across existing state of the art methods. Mentions of a detail across accounts are exponentially distributed with context-invariant details appearing dozens to hundreds of times more than context-dependent details.[^9] Furthermore, crisis start and stop dates are arbitrary, and the historical record points to many precursor events as necessary detail for understanding later events. Figure 2 compares ICBe's recall with that of existing datasets for the two case studies detailed in Section 4. ICBe strictly dominates all of the systems but ICEWs in recall though we note that the small sample sizes mean these systems should be considered statistically indistinguishable. Across all existing datasets and ICBe, recall increases with the number of document mentions which is an important sign of validity for both them and our benchmark. The one outlier is Phoenix which in the Cuban Missile Crisis case is so noisy that it's recall curve is flat to decreasing as mentions increase. The two episode-level datasets (MIDs and ICM) have low coverage of contextual details. The two other dictionary systems ICEWs and Terrier have higher coverage, with ICEWs outperforming Terrier. Importantly our corpus of ICB narratives has high recall of frequently mentioned details giving us confidence in how those summaries were constructed, and ICBe lags only slightly behind showing that it left little additional information on the table.

@@

@@

Figure 3 shows the location of every sentence from the ICBe corpus in semantic space as embedded using the same large language model as before, and the median location of each ICBe event tag applied to those sentences.[^10] Labels reflect the individual leaves of the ontology and colors reflect the higher level coarse branch nodes of the ontology. If ICBe has high precision, substantively similar tags ought to have been applied to substantively similar source text, which is what we see both in two dimensions in the main plot and via hierarchical clustering on all dimensions in the dendrogram along the right-hand side.[^11]

[^10]: We preprocess sentences to replace named entities with a generic Entity token.

[^11]: Hierarchical clustering on cosine similarity and with Ward's method.

@@

We release several open-source products along with supporting code and documentation to further advance the study of international relations, event extraction, and natural language processing. The first is the International Crisis Behavior Events (ICBe) dataset, an event-level aggregation of what took place during the crises identified by the ICB project. These data are appropriate for statistical analysis of hard questions about the sequencing of events (e.g. escalation and de-escalation of conflicts).[^12] Second, we provide a coder-level disaggregation with multiple codings of each sentence by experts and undergrads that allows for the introduction of uncertainty and human interpretation of events. Further, we release a direct mapping from the codings to the source text at the sentence level as a new resource for natural language processing. Finally, we provide a companion website that incorporates detailed visualizations of all of the data introduced here at crisisevents.org.

[^12]: Using ICBe data, @gannonOneIfLand2022 finds that cross-domain crises are shorter and less violent than same-domain crises.

>>>>

We find substantive variation in recall across existing state of the art methods. Mentions of a detail across accounts are exponentially distributed with context-invariant details appearing dozens to hundreds of times more than context-dependent details.[^9] Furthermore, crisis start and stop dates are arbitrary, and the historical record points to many precursor events as necessary detail for understanding later events. Figure 2 compares ICBe's recall with that of existing datasets for the two case studies detailed in Section 4. ICBe strictly dominates all of the systems but ICEWs in recall though we note that the small sample sizes mean these systems should be considered statistically indistinguishable. Across all existing datasets and ICBe, recall increases with the number of document mentions which is an important sign of validity for both them and our benchmark. The one outlier is Phoenix which in the Cuban Missile Crisis case is so noisy that it's recall curve is flat to decreasing as mentions increase. The two episode-level datasets (MIDs and ICM) have low coverage of contextual details. The two other dictionary systems ICEWs and Terrier have higher coverage, with ICEWs outperforming Terrier. Importantly our corpus of ICB narratives has high recall of frequently mentioned details giving us confidence in how those summaries were constructed, and ICBe lags only slightly behind showing that it left little additional information on the table.[^10]

@@

[^10]: Although Figure 2 focuses only on two crises, the synthetic narrative approach and recall comparison can, and should, be more broadly applied to all international crises in a way that could reveal systematic blindspots across datasets.

@@

Figure 3 shows the location of every sentence from the ICBe corpus in semantic space as embedded using the same large language model as before, and the median location of each ICBe event tag applied to those sentences.[^11] Labels reflect the individual leaves of the ontology and colors reflect the higher level coarse branch nodes of the ontology. If ICBe has high precision, substantively similar tags ought to have been applied to substantively similar source text, which is what we see both in two dimensions in the main plot and via hierarchical clustering on all dimensions in the dendrogram along the right-hand side.[^12]

[^11]: We preprocess sentences to replace named entities with a generic Entity token.

[^12]: Hierarchical clustering on cosine similarity and with Ward's method.

@@

We release several open-source products along with supporting code and documentation to further advance the study of international relations, event extraction, and natural language processing. The first is the International Crisis Behavior Events (ICBe) dataset, an event-level aggregation of what took place during the crises identified by the ICB project. These data are appropriate for statistical analysis of hard questions about the sequencing of events (e.g. escalation and de-escalation of conflicts).[^13] Second, we provide a coder-level disaggregation with multiple codings of each sentence by experts and undergrads that allows for the introduction of uncertainty and human interpretation of events. Further, we release a direct mapping from the codings to the source text at the sentence level as a new resource for natural language processing. Finally, we provide a companion website that incorporates detailed visualizations of all of the data introduced here at crisisevents.org.

[^13]: Using ICBe data, @gannonOneIfLand2022 finds that cross-domain crises are shorter and less violent than same-domain crises.

### 2023-11-20 replication_paper/psrm_draft/ICBe_manuscript.tex
fig 2 footnote future research

We find substantive variation in recall across existing state of the art
methods. Mentions of a detail across accounts are exponentially
distributed with context-invariant details appearing dozens to hundreds
of times more than context-dependent details.\footnote{As the ICB
  narratives are intended to explain conflictual behavior in a political
  context, many of the missing events concern more economic components
  of conflict (eg. nationalizing a foreign business). Even when they
  occur in the context of a crisis, these events largely fall outside
  the sample of information on which ICBe's ontology is currently
  trained. Even with this limitation, ICBe is more comprehensive than
  the existing datasets that do try to code the economic dimensions of
  these crises. We see expanding the ontology to broader international
  phenomenon as a promising future implementation of our model.}
Furthermore, crisis start and stop dates are arbitrary, and the
historical record points to many precursor events as necessary detail
for understanding later events. Figure 2 compares ICBe's recall with
that of existing datasets for the two case studies detailed in Section
4. ICBe strictly dominates all of the systems but ICEWs in recall though
we note that the small sample sizes mean these systems should be
considered statistically indistinguishable. Across all existing datasets
and ICBe, recall increases with the number of document mentions which is
an important sign of validity for both them and our benchmark. The one
outlier is Phoenix which in the Cuban Missile Crisis case is so noisy
that it's recall curve is flat to decreasing as mentions increase. The
two episode-level datasets (MIDs and ICM) have low coverage of
contextual details. The two other dictionary systems ICEWs and Terrier
have higher coverage, with ICEWs outperforming Terrier. Importantly our
corpus of ICB narratives has high recall of frequently mentioned details
giving us confidence in how those summaries were constructed, and ICBe
lags only slightly behind showing that it left little additional
information on the table.

>>>>

We find substantive variation in recall across existing state of the art
methods. Mentions of a detail across accounts are exponentially
distributed with context-invariant details appearing dozens to hundreds
of times more than context-dependent details.\footnote{As the ICB
  narratives are intended to explain conflictual behavior in a political
  context, many of the missing events concern more economic components
  of conflict (eg. nationalizing a foreign business). Even when they
  occur in the context of a crisis, these events largely fall outside
  the sample of information on which ICBe's ontology is currently
  trained. Even with this limitation, ICBe is more comprehensive than
  the existing datasets that do try to code the economic dimensions of
  these crises. We see expanding the ontology to broader international
  phenomenon as a promising future implementation of our model.}
Furthermore, crisis start and stop dates are arbitrary, and the
historical record points to many precursor events as necessary detail
for understanding later events. Figure 2 compares ICBe's recall with
that of existing datasets for the two case studies detailed in Section
4. ICBe strictly dominates all of the systems but ICEWs in recall though
we note that the small sample sizes mean these systems should be
considered statistically indistinguishable. Across all existing datasets
and ICBe, recall increases with the number of document mentions which is
an important sign of validity for both them and our benchmark. The one
outlier is Phoenix which in the Cuban Missile Crisis case is so noisy
that it's recall curve is flat to decreasing as mentions increase. The
two episode-level datasets (MIDs and ICM) have low coverage of
contextual details. The two other dictionary systems ICEWs and Terrier
have higher coverage, with ICEWs outperforming Terrier. Importantly our
corpus of ICB narratives has high recall of frequently mentioned details
giving us confidence in how those summaries were constructed, and ICBe
lags only slightly behind showing that it left little additional
information on the table.\footnote{Although Figure 2 focuses only on two
  crises, the synthetic narrative approach and recall comparison can,
  and should, be more broadly applied to all international crises in a
  way that could reveal systematic blindspots across datasets.}
